Diffusion local times and the time until recombination

### 2015-10-15 notes/diffusion-local-times.Rmd
more comparisons

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.align='center')
```

*Useful functions used below:*
```{r setup_fns}
pcline <- function (x,r,tau=1,sigma=1) {
    (1/2)*( 1 - sign(x)*zeta(x=x/sqrt(tau*sigma^2),r=r*tau) )
}
tanh_cline <- function (x,s,sigma=1) {
    (1/2)*(1+tanh(-2*x*sqrt(s)/sigma))
}
tanh_drift <- function (x,s,sigma=1) {
    # log-derivative of tanh_cline
    -(2*sqrt(s)/sigma) * (1-tanh(-2*x*sqrt(s)/sigma)^2) / (1+tanh(-2*x*sqrt(s)/sigma))
}
tanh_het <- function (x,s,sigma=1) {
    # prob of being in a het = (1-s)(1-p)/(1-s(1-p))
    pp <- tanh_cline(x,s,sigma=sigma)
    (1-s)*(1-pp)/(1-s*(1-pp))
}
sim_diffusion <- function (drift,tau,nwalks,ntimes,xinit=0,...) {
    # want total time to be tau, so:
    dt <- tau/ntimes
    # put the noise in rws
    rws <- matrix( c( rep_len(xinit,nwalks), # initial conditions
                 rnorm(nwalks*(ntimes-1)) ), nrow=ntimes, ncol=nwalks, byrow=TRUE )
    # and then integrate over it
    for (k in 2:ntimes) {
        rws[k,] <- rws[k-1,]+sqrt(dt)*rws[k,]+dt*drift(rws[k-1,],...)
    }
    attr(rws,"dt") <- dt
    return(rws)
}
```

@@

Bazykin's approximation to a cline is
$$
p(x) = \frac{1}{2} (1+\tanh(-2x\sqrt{s}/\sigma))
$$
Here's what that looks like:
```{r tanh_cline}
s <- 0.1
tanh_cline
xx <- seq(-15,15,length.out=500)
plot(xx, (-1)*tanh_drift(xx,s=s), type='l', col='red', xlab="space", ylab='proportion or drift')
lines(xx, tanh_cline(xx,s=s), col='black')
lines(xx, 2*tanh_cline(xx,s=s)*(1-tanh_cline(xx,s=s)), col='blue')
lines(xx, tanh_het(xx,s=s), col='green')
legend("topleft",lty=1,col=c("black","red","blue","green"),
        legend=c("p(x)","-log(p(x))'","2p(1-p)","prob of het"))
```

@@

```{r sim_general, fig.height=3*fig.dim}
weighted_time <- function (rws, weight.fn, ...) {
    colSums( weight.fn(rws,...) ) * attr(rws,"dt")
}
nwalks <- 1e4
ntimes <- 1e3
s <- 0.1
tau <- 10
lt.comparison <- list(
        tanh=weighted_time( 
               sim_diffusion( tanh_drift,tau,nwalks,ntimes,s=s ),
               tanh_het, s=s
           ),
        reflected=weighted_time(
                sim_diffusion( function(...){0}, tau, nwalks, ntimes ),
                function (x,s) { ifelse(abs(x)<1/(2*sqrt(s)), 1/2, 0) }, s=s
            ),
        reflected_tanh=weighted_time(
                sim_diffusion( function(...){0}, tau, nwalks, ntimes ),
                tanh_het, s=s
            ),
        pushed=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s ),
                function (x) { x>0 }
            ),
        pushed_tanh=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s ),
                tanh_het, s=s
            )
        )
layout(seq_along(lt.comparison))
for (k in seq_along(lt.comparison)) {
    hist(lt.comparison[[k]],xlim=range(lt.comparison),main=names(lt.comparison)[k],breaks=50,xlab="time in hets")
}
```

>>>>

```{r doc_setup, include=FALSE}
fig.dim <- 5
knitr::opts_chunk$set(fig.width=2*fig.dim,fig.height=fig.dim,fig.align='center')
```

*Useful functions used below:*
```{r setup_fns}
pcline <- function (x,r,tau=1,sigma=1) {
    (1/2)*( 1 - sign(x)*zeta(x=x/sqrt(tau*sigma^2),r=r*tau) )
}
tanh_cline <- function (x,s,sigma=1) {
    (1/2)*(1+tanh(-2*x*sqrt(s)/sigma))
}
tanh_drift <- function (x,s,sigma=1) {
    # log-derivative of tanh_cline
    -(2*sqrt(s)/sigma) * (1-tanh(-2*x*sqrt(s)/sigma)^2) / (1+tanh(-2*x*sqrt(s)/sigma))
}
tanh_het <- function (x,s,sigma=1) {
    # prob of being in a het = (1-s)(1-p)/(1-s(1-p))
    pp <- tanh_cline(x,s,sigma=sigma)
    (1-s)*(1-pp)/(1-s*(1-pp))
}
sim_diffusion <- function (drift,tau,nwalks,ntimes,xinit=0,...) {
    # want total time to be tau, so:
    dt <- tau/ntimes
    # put the noise in rws
    rws <- matrix( c( rep_len(xinit,nwalks), # initial conditions
                 rnorm(nwalks*(ntimes-1)) ), nrow=ntimes, ncol=nwalks, byrow=TRUE )
    # and then integrate over it
    for (k in 2:ntimes) {
        rws[k,] <- rws[k-1,]+sqrt(dt)*rws[k,]+dt*drift(rws[k-1,],...)
    }
    attr(rws,"dt") <- dt
    return(rws)
}
weighted_time <- function (rws, weight.fn, ...) {
    colSums( weight.fn(rws,...) ) * attr(rws,"dt")
}
```

@@

Bazykin's approximation to a cline is
$$
p(x) = \frac{1}{2} (1+\tanh(-2x\sqrt{s}/\sigma))
$$
Here's what that looks like:
```{r tanh_cline}
tanh_cline
s <- 0.1
xx <- seq(-15,15,length.out=500)
plot(xx, (-1)*tanh_drift(xx,s=s), type='l', col='red', xlab="space", ylab='proportion or drift')
lines(xx, tanh_cline(xx,s=s), col='black')
lines(xx, 2*tanh_cline(xx,s=s)*(1-tanh_cline(xx,s=s)), col='blue')
lines(xx, tanh_het(xx,s=s), col='green')
legend("topleft",lty=1,col=c("black","red","blue","green"),
        legend=c("p(x)","-log(p(x))'","2p(1-p)","prob of het"))
```

@@

Here are distributions of $T=\int_0^{`r tau`} f_H(X_t) dt$
for two different starting points: $x=0$ and $x=-3$,
and five different models:

- **tanh:** the ``correct'' model;
- **reflected:** a Brownian motion reflected at zero and counting one-half time in a neighborhood of size $\sigma/\sqrt{s}$;
- **reflected tanh:** a Brownian motion reflected at zero and counting time weighted by the $\tanh$ cline;
- **pushed:** a Brownian motion pushed back at rate $4\sigma\sqrt{s}$ above zero and counting time above zero;
- **pushed tanh:** a Brownian motion pushed back at rate $4\sigma\sqrt{s}$ above zero and counting time weighted by the $\tanh$ cline.

```{r sim_general, fig.height=3*fig.dim, cache=TRUE}
nwalks <- 1e4
ntimes <- 1e3
s <- 0.1
tau <- 10
initvals <- c(0,-3)
lt.comparison <- lapply( initvals, function (xinit) { list(
        tanh=weighted_time( 
               sim_diffusion( tanh_drift,tau,nwalks,ntimes,s=s, xinit=xinit ),
               tanh_het, s=s
           ),
        reflected=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                function (x,s) { ifelse(abs(x)<1/(2*sqrt(s)), 1/2, 0) }, s=s
            ),
        reflected_tanh=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                tanh_het, s=s
            ),
        pushed=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                function (x) { x>0 }
            ),
        pushed_tanh=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                tanh_het, s=s
            )
        ) } )
layout(matrix(1:(length(lt.comparison)*length(lt.comparison[[1]])),ncol=2))
for (k in seq_along(lt.comparison)) {
    for (j in seq_along(lt.comparison[[k]])) {
        hist(lt.comparison[[k]][[j]],xlim=range(lt.comparison),main=names(lt.comparison[[k]])[j],breaks=seq(0,tau,length.out=50),xlab="time in hets")
    }
}
```

Of course, we aren't actually trying to estimate the distribution of $T$ itself,
but rather the probabilities of recombination (the Laplace transform).
Here are these probabilities, as a function of $r$
(the second plots are on a log-log scale):
```{r recomb_prob_comparisons, cache=TRUE, depends="sim_general"}
rvals <- exp(seq(log(.001),log(0.5),length.out=400))
precombs <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) sapply( rvals, function (r) { mean((-1)*expm1(-r*lt)) } ) )
    } )
layout(t(1:2))
for (k in seq_along(precombs)) {
    matplot( rvals, precombs[[k]], type='l', col=1:ncol(precombs[[k]]), main=paste("x =",initvals[k]), ylim=range(precombs), ylab="prob of recomb", xlab='r', lty=1 )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
for (k in seq_along(precombs)) {
    matplot( rvals, precombs[[k]], type='l', col=1:ncol(precombs[[k]]), log='xy', main=paste("x =",initvals[k]), ylim=range(precombs), ylab="prob of recomb", xlab='r', lty=1  )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
```

Here's the same thing for a much longer time:
```{r sim_general_longer, fig.height=3*fig.dim, cache=TRUE}
nwalks <- 1e3
ntimes <- 1e4
s <- 0.1
tau <- 100
initvals <- c(0,-3)
lt.comparison <- lapply( initvals, function (xinit) { list(
        tanh=weighted_time( 
               sim_diffusion( tanh_drift,tau,nwalks,ntimes,s=s, xinit=xinit ),
               tanh_het, s=s
           ),
        reflected=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                function (x,s) { ifelse(abs(x)<1/(2*sqrt(s)), 1/2, 0) }, s=s
            ),
        reflected_tanh=weighted_time(
                -abs(sim_diffusion( function(...){0}, tau, nwalks, ntimes, xinit=xinit )),
                tanh_het, s=s
            ),
        pushed=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                function (x) { x>0 }
            ),
        pushed_tanh=weighted_time(
                sim_diffusion( function(x,s){ifelse(x>0,-4*sqrt(s),0)}, tau, nwalks, ntimes, s=s, xinit=xinit ),
                tanh_het, s=s
            )
        ) } )
tbreaks <- seq(0,tau,length.out=50)
Tdists <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) tabulate(findInterval(lt,tbreaks),nbins=length(tbreaks)-1) )
    } )
rvals <- exp(seq(log(.001),log(0.5),length.out=400))
precombs <- lapply( lt.comparison, function (lts) {
        sapply( lts, function (lt) sapply( rvals, function (r) { mean((-1)*expm1(-r*lt)) } ) )
    } )
layout(t(1:2))
for (k in seq_along(precombs)) {
    matplot( tbreaks[-1]-diff(tbreaks)/2, Tdists[[k]], type='l', col=1:ncol(Tdists[[k]]), main=paste("x =",initvals[k]), ylim=range(Tdists), xlab="time in hets", lty=1 )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
for (k in seq_along(precombs)) {
    matplot( rvals, precombs[[k]], type='l', col=1:ncol(precombs[[k]]), main=paste("x =",initvals[k]), ylim=range(precombs), ylab="prob of recomb", xlab='r', lty=1 )
    legend("topleft",lty=1,col=1:ncol(precombs[[k]]),legend=colnames(precombs[[k]]))
}
```
